
%   COMP3310 DV routing tutorial
%   Written by Hugh Fisher, ANU, 2024-
%   Creative Commons CC-BY-NC-SA-4.0 license

%   LaTeX markup for XeTeX with a standard texlive distribution
%   Command to build:
%       xelatex 0N-routing.tex

\input{../common/formatting}

\begin{document}

\TITLE{Distance Vector Routing}

\MINOR{Outline}

In this tutorial you will

\DOT See how distance vector routers self-configure

\DOT See how routers adapt \DSH or not \DSH to changes

\DOT Try sending bad information to routers

\begin{IMPORTANT}
The instructions for this exercise will have even less detail
than usual. Routing is inherently dynamic, unpredictable, and
messy; so the most likely answer to any questions you have is
\DQ{try it and see what happens}.
\end{IMPORTANT}


\SECTION{Setup}

This exercise is designed for the Computer Science Linux labs.
It is not a stand alone simulator that runs on one machine.
Instead this is designed for a LAN with multiple PCs,
each of which becomes a \SQ{router}.

Ideally there are at least four people. You should be sitting
close together so you can see what is happening on
other screens.

If there are less than four, you can log on to more than one
PC in a lab and run the same program on each. On all but one
PC, use \texttt{Ctrl-Alt-F1} before logging on to get
a plain text console, no GUI. (For some reason the keyboard
may stop working on other PCs if you have multiple GUI sessions
open.) After logging out, \texttt{Ctrl-Alt-F7} to restore
the GUI.

These programs are Python or Java using just the standard
library so could run elsewhere. All of the hosts would need
to be connected to a single Ethernet LAN or a single WiFi access
point, with multicast. Because of this it probably won't work
on the ANU campus WiFi.

You don't need \NAME{WireShark} or \NAME{tcpdump}, as these
\SQ{routers} are not using a standard protocol and will print
out what is happening for you.

\SECTION{Running a router}

Download and unzip the tutorial exercise file. As usual, these
are CLI programs so open a shell and \texttt{cd} into the new
directory. (If you are logged onto more than one PC, do the
same on the others.) You will find two sub directories,
\texttt{PyDV} for the Python version and \texttt{JDV} for Java,
a few text files, and a copy of these instructions.

\textbf{Starting and stopping}

The routing simulator has been written as a \textbf{package}
(both Python and Java), so is run slightly differently.
You need to be in the top level directory, not the directory
containing the code.

For the Python version:
\CODELINE{python -m PyDV.dvRouter}

For the Java version, compile first and then run
\begin{CODE}\begin{verbatim}
javac JDV/*.java
java JDV.DVRouter
\end{verbatim}\end{CODE}

Note that there is no configuration, no coordination with anyone
else. Ideally a router can bootstrap / auto configure / self
configure itself every time it is switched on and never need
human intervention.

This program will start printing out messages. (Real routers
are usually in a rack somewhere and nobody looks at them.)
It will keep running until you \texttt{Ctrl-C}, and again you
don't need to tell anyone else that you are doing so.

Stop your router and read on.

\textbf{Options}

By default the router program prints out quite a lot about
what is happening and what it is doing. After your first
few runs you can use

\CODELINE{python -m PyDV.dvRouter -q}
or
\CODELINE{java JDV.DVRouter -q}

which means \DQ{shut up and just show me the routing tables}.

There's also a \texttt{-debug} option which will print out
enormous amounts of detail, and a few others. \texttt{-help}
will list them all.


\SECTION{Domains and links}

Each Computer Science lab is a single subnet, so the PCs
can talk directly to each other, no routing required.
For this exercise each router program will simulate being a
\NAME{domain}, an isolated network, that is connected to
other routers by point to point \NAME{links}.

\textbf{Domains}

When the router program starts it prints out its own
router name, usually the name for its own IP address;
and its own domain, an imaginary
subnet / area / zone that needs a router.

The domain name is randomly picked from a text file
(\texttt{domains.txt}), but the router program on the same
PC will always pick the same domain. This is so you can stop
and restart your router during the simulation.

It is OK if more than one router picks the same domain.
Real networks often have multiple routers for redundancy.

\textbf{Links}

A router is the gateway between one collection of hosts that
can all talk to each locally, called a \NAME{domain} for this
exercise; and the routers for other domains. The connections
between routers are often called \NAME{links} but they are
really just another network cable.

Real routers detect their links by counting the plugged in
Ethernet (or whatever) cables and testing if they connect.
For this exercise the programs send messages back and forth,
until after ten to thirty seconds every router will have at
least one \SQ{link}. They won't have links to every other router,
just a few, so some domains will be more than one network
hop away if there are many routers.

\begin{IMPORTANT}
The \SQ{routers} have to be on the same LAN to create simulated
links, but the actual routing protocol runs entirely over point
to point (TCP) connections. There is no broadcast / multicast
cheating going on underneath.
\end{IMPORTANT}


\SECTION{Route formation}

After all that explanation, it's finally time to do some routing.

\STEP
A routing exercise needs multiple routers, but the amount of
output to read will keep growing. If there are more than about
eight, it may be best to divide into smaller groups by using
different link multicast addresses for each.

The default group uses address \texttt{224.0.0.70}. You can change
this by running the program and specifying a different address, eg

\CODELINE{\ldots -mcast 224.0.0.71}

Only routers using the same multicast address will form links
to each other.


\STEP
About three-quarters of the people in your group should start their
routers at more or less the same time. Don't use \texttt{-q}, the
more detailed output will be useful.

A router starts by printing out its own name and domain, and then
the initial routing table. This should look something like:

\begin{CODE}\begin{verbatim}
Router: nanotech Domain: ENCOM
...
Routing table
  ENCOM                   : nanotech        :  0
\end{verbatim}\end{CODE}

A distance vector router starts up knowing only who it is and
its own domain, which is zero network hops away.


\STEP
The routers will establish links, and start receiving cost
tables from other routers. Each \SQ{beat} the routing table gets
updated based on the new costs from other routers. Start counting
the beats for your router and how many changes are made each time.

The router prints out its entire routing table, which will usually
have multiple entries for each domain, sorted by cost. The first
entry, the lowest cost, is what would be used for actual traffic.

After a while the routers should all stablise, with no changes to
the routing tables.

Of the first entries for the domains, what is the longest
(highest cost) path in the network?

Are the routes symmetrical? Meaning, if router A wants to
reach the domain for router B, is its cost the same as the cost
for router B to reach the domain for router A?


\STEP
Start the remaining router(s) in the group.

How many beats does it take for all the routing tables to
become stable again?

Do some routers stabilise faster than others?


\STEP
\texttt{Ctrl-C} one router and watch what happens to the others.

How long does it take for other routers to notice?

Are there any \DQ{count to infinity} loops? These will
show up as two or more routers steadily increasing the cost to
a domain by +1 each beat.


\STEP
Start the router again. Watch what happens on the other routers.


\STEP
Shut down all the routers.


\SECTION{Misinformation}

For this section you can use the \texttt{-q} option on some or
all of the routers to simplify the output.

With all the routers shut down, pick one router to be \SQ{evil}
and one other domain as the \SQ{target}.

Start all the non-evil routers and wait for them to stabilise.

Meanwhile, on the evil router, open file \texttt{evil.routes}.
Add a line

\CODELINE{\textit{target-domain} : -4}

Start the evil router with an extra CLI option

\CODELINE{\ldots -q -evil evil.routes}

The first route table printed will now have two entries, not one.

Watch what happens as this spreads across the links.

\begin{IMPORTANT}
This really happened to the early Internet, by malfunction rather
than malicious. A \DQ{black hole} router attracts packets to it
which then vanish.

It is still a problem today. Organisations and
countries have to receive routing messages from other organisations
and countries, and they might not trust them all.
\end{IMPORTANT}


\SECTION{Optional}


\textbf{Poison Reverse}

When a link to another router is lost, a distance vector router
can respond in a few different ways. The program in this exercise
just forgets about the route.

The code that controls this is in
method \texttt{drop}, in \texttt{dvRouter.py} or \texttt{DVRouter.java}.
There are two possibilities,
one of which needs to be uncommented. Make the change, study
the code, and run the route formation exercise again to see
what happens when routers disappear.


\textbf{Persistence}

Every time the router restarts it must rebuild the routing table
based on messages as they are received, which can take a long time.
(By computer standards, twenty seconds is a very long time.)

Modify the router so that at the end of
each beat it writes the current routing table out to a file,
perhaps named \texttt{cache.routes}. When the router starts to
run it should check if this file exists and if so load all the
entries from it. (Use the method \texttt{addFixedRoutes} as a
starting point, with a third router name field in each line.)

Persistent configuration helps a lot when a router crashes due
to a power blip and restarts a few seconds later. It's not so
helpful if the router is shut down, shipped to another continent,
and then switched on again.

%%

\COPYRIGHT

\end{document}
